---
title: "Fit MOSM GP to 2-band synthetic example"
format: 
  html:
    embed-resources: true
---

This notebook contains the workflow for:

- loading groundtruth 2-band (MOGP) time-series, 
- fitting a MOGP model, 
- checking which posterior draws produce valid covariance matrices, and 
- visualing the posteriors of the valid draws.

```{r}
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cmdstanr)
library(posterior)
library(bayesplot)
library(readr)
library(glue)
library(tictoc)

register_knitr_engine(override = FALSE)

theme_set(theme_bw())
color_scheme_set("viridis") # bayesplot

source(here("mogp-utils.R"))

filename_stub <- "D2-Q1-seed6893-n200-mnar0.3-sigma0.1"

EXECUTE_SAMPLING <- FALSE
SAMPLING_SEED <- ifelse(EXECUTE_SAMPLING, round(runif(n = 1, 0, 999)), 119)
SAMPLING_SEED
```

```{r}
D <- 2
Q <- 1

input_df <- read_csv(
  here("data", glue("{filename_stub}-sim.csv")), 
  show_col_types = FALSE)

input_params <- read_tsv(
  here("data", glue("{filename_stub}-param.tsv")), 
  show_col_types = FALSE)

input_df |> 
  ggplot() +
  aes(x = t, y = y, colour = factor(d)) +
  geom_linerange(aes(ymax = y + y_se, ymin = y - y_se), colour = "black") +
  geom_point() +
  geom_line(aes(y = f), alpha = 0.8) +
  #facet_grid(rows = vars(d), scales = "fixed") +
  labs(colour = "Band")

ns <- table(input_df$d)
```

```{r}
input_params
```


# Model Fitting

```{r}
mosmgp_zerophi_stan <- cmdstan_model(here("stan", "moskgp.stan"))
```

```{r}
SAMPLING_SEED
```

## Sampling

```{r}
data_list <- list(
  D = D,
  N = NROW(input_df),
  ns = ns,
  d = input_df$d,
  x = input_df$t,
  y = input_df$y,
  y_se = input_df$y_se
)
```

```{r}
format(start_time <- Sys.time())
```

```{r}
if (EXECUTE_SAMPLING) {
  
  output <- mosmgp_zerophi_stan$sample(
    data = data_list,
    seed = SAMPLING_SEED,
    chains = 6,
    parallel_chains = 6,
    iter_warmup = 1000,
    iter_sampling = 1000,
    refresh = 500,
    show_exceptions = FALSE,
    show_messages = TRUE
  )
  
  output$save_object(
    here("RDS", glue("{filename_stub}_mogp.RDS")))
  
} else {
  output <- readRDS(
    here("RDS", glue("{filename_stub}_mogp.RDS")))
}
```

```{r}
format(finish_time <- Sys.time())
difftime(finish_time, start_time)  
```

# Results

```{r}
vars <- c("w", "Sigma", "mu", "theta", "phi", "deltaTheta")
output$summary(variables = vars, 
               posterior::default_convergence_measures())
```

```{r}
mcmc_trace(output$draws(vars))
```


```{r}
valid_chains <- c(3:6)
output_draws <- output$draws(variables = vars, format = "df") |> 
  subset_draws(chain = valid_chains)

output_draws_mat <- as.matrix(output_draws)

color_scheme_set("mix-red-blue")
output_draws |> 
  summary(posterior::default_convergence_measures())
```

```{r}
output_draws |> mcmc_trace()
```

```{r}
output$draws(variables = c("theta")) |> mcmc_pairs()
```

## Posterior Predictive Check

### Target locations

```{r}
n_star <- 100 # in each band
N_star <- n_star * D
ns_star = rep(n_star, D)

star_df <- data.frame(
  t_star = rep(seq(from = 0, to = 1, length.out = n_star), D),
  d_star = as.integer(rep(1:D, each = n_star))
)
```

```{r}
# Draw posterior predictive samples using the posterior means
posterior_means <- as.data.frame(output_draws) |> 
  select(`w[1]`:deltaTheta) |> 
  colMeans()
posterior_means
```

```{r}
input_params
```


```{r eval=FALSE}
mean_ws <- posterior_means[1:2]
mean_Sigmas <- posterior_means[3:4]
mean_mus <- posterior_means[5:6]
mean_thetas <- posterior_means[7:8]
mean_phis <- posterior_means[9:10]

n_draws <- 10

mean_draws <- postpred_Q1_draws(
  n_draws = n_draws, 
  xs = input_df$t, ys = input_df$y, ys_se = input_df$y_se, ds = input_df$d, ns = ns,  
  xs_star = star_df$t_star, ds_star = star_df$d_star, ns_star = ns_star, 
  D = D, 
  w = mean_ws, Sigma = mean_Sigmas, mu = mean_mus, theta = mean_thetas, phi = mean_phis
)

ggplot() +
  geom_line(data = mean_draws, 
            aes(x = x_star, y = f_star, 
                colour = d_star,
                group = interaction(d_star,.draw)), alpha = 0.4) +
  geom_point(data = input_df, aes(x = t, y = y, colour = factor(d)), size = 1) 
```

```{r eval=FALSE}
# Draw posterior predictive samples using a particular posterior draw
r <- 100
this_w <- output_draws_mat[r, 1:2]
this_Sigma <- output_draws_mat[r, 3:4]
this_mu <- output_draws_mat[r, 5:6]
this_theta <- output_draws_mat[r, 7:8]
this_phi <- output_draws_mat[r, 9:10]

pp_draws <- postpred_Q1_draws(
  n_draws = 10,
  xs = input_df$t, ys = input_df$y, ys_se = input_df$y_se, ds = input_df$d, ns = ns,  
  xs_star = star_df$t_star, ds_star = star_df$d_star, ns_star = ns_star, 
  D = D, 
  w = this_w, Sigma = this_Sigma, mu = this_mu, theta = this_theta, phi = this_phi
)

ggplot() +
  geom_line(data = pp_draws, 
            aes(x = x_star, y = f_star, 
                colour = d_star,
                group = interaction(d_star,.draw)), alpha = 0.4) +
  geom_point(data = input_df, aes(x = t, y = y, colour = factor(d)), size = 1) 
```


```{r}
if (EXECUTE_SAMPLING) {
  
  tic()
  outputs <- check_postpred_draws_validity(
    xs = input_df$t, ys = input_df$y, ys_se = input_df$y_se, ds = input_df$d, ns = ns, 
    xs_star = star_df$t_star, ds_star = star_df$d_star, ns_star = ns_star, 
    D = D, 
    ws = output_draws_mat[,1:2],
    Sigmas = output_draws_mat[,3:4],
    mus = output_draws_mat[,5:6],
    thetas = output_draws_mat[,7:8],
    phis = output_draws_mat[,9:10]
  )
  toc()
  
  write_rds(outputs, here("RDS", glue("{filename_stub}_draws{SAMPLING_SEED}_valid.RDS")))
} else {
  
  outputs <- read_rds(here("RDS", glue("{filename_stub}_draws{SAMPLING_SEED}_valid.RDS")))
}

```

### Test code for faster validity checking


```{r eval=FALSE}
tic()
valid_draws_checks <- batch_check_K_validity(
  xs = input_df$t, xs_star = star_df$t_star, 
  ds = input_df$d, ds_star = star_df$d_star, 
  ns = ns, ns_star = ns_star,
  D = D,
  ws = output_draws_mat[,1:2],
  Sigmas = output_draws_mat[,3:4],
  mus = output_draws_mat[,5:6],
  thetas = output_draws_mat[,7:8],
  phis = output_draws_mat[,9:10] # matrix(0, nrow = NROW(output_draws_mat), ncol = D) 
)
toc()
```

```{r}
valid_draws_idx <- which(outputs[[2]] == "Warning") #which(valid_draws_checks == "Warning")
valid_draws_df <- output_draws[valid_draws_idx,]
valid_draws_mat <- as.matrix(valid_draws_df)
```

```{r eval=FALSE}
tic()
valid_postpred_draws_df <- postpred_from_valid_draws(
  xs = input_df$t, ys = input_df$y, ys_se = input_df$y_se, 
  ds = input_df$d, ns = ns, 
  xs_star = star_df$t_star, 
  ds_star = star_df$d_star, ns_star = ns_star,
  D = D,
  ws = output_draws_mat[,1:2],
  Sigmas = output_draws_mat[,3:4],
  mus = output_draws_mat[,5:6],
  thetas = output_draws_mat[,7:8],
  phis = matrix(0, nrow = NROW(output_draws_mat), ncol = D) #output_draws_mat[,9:10]
)
toc()
```


```{r eval=FALSE}
ggplot() +
  geom_line(data = valid_postpred_draws_df, aes(x = x_star, y = f_star, 
      colour = d_star,
      group = interaction(.draw,d_star))) +
  geom_point(data = input_df, aes(x = t, y = y), size = 1) 

```

## Hyperparameter Posteriors

```{r}
estimated_params <- sapply(select(valid_draws_df, `w[1]`:deltaTheta), mean)
estimated_params
```

```{r}
# Function estimates the mode of a sample
estimate_mode <- function(x) {
  d <- density(x, n = 2^10)
  d$x[which.max(d$y)]
}

sapply(select(valid_draws_df, `w[1]`:deltaTheta), estimate_mode)
```

```{r}
input_params
```

```{r}
diff(input_params$theta)
estimated_params['deltaTheta'] / pi
```

```{r}
valid_draws_df |> 
  select(.draw, starts_with("w")) |> 
  pivot_longer(2:3) |> 
  drop_na(value) |> 
  ggplot() +
  aes(x = value) +
  geom_histogram() +
  facet_wrap(vars(name), scales = "free_y")
```

```{r}
valid_draws_df |> 
  select(.draw, starts_with("Sigma")) |> 
  pivot_longer(2:3) |> 
  drop_na(value) |> 
  ggplot() +
  aes(x = value) +
  geom_histogram() +
  facet_wrap(vars(name), scales = "free_y")
```

```{r}
valid_draws_df |> 
  select(.draw, starts_with("mu")) |> 
  pivot_longer(2:3) |> 
  drop_na(value) |> 
  ggplot() +
  aes(x = value) +
  geom_histogram() +
  facet_wrap(vars(name), scales = "free")
```

```{r}
valid_draws_df |> 
  select(.draw, starts_with("theta")) |> 
  pivot_longer(2:3) |> 
  drop_na(value) |> 
  ggplot() +
  aes(x = value) +
  geom_histogram() +
  facet_wrap(vars(name), scales = "free_y")
```

```{r eval=FALSE}
valid_draws_df |>
  select(.draw, starts_with("phi")) |>
  pivot_longer(2:3) |>
  drop_na(value) |>
  ggplot() +
  aes(x = value) +
  geom_histogram() +
  facet_wrap(vars(name), scales = "free_y")
```

```{r}
valid_draws_df |> 
  pivot_longer(cols = c(`theta[1]`, `theta[2]`), names_to = "Parameter") |> 
  ggplot() +
  aes(x = value, fill = Parameter) +
  geom_histogram(position = "identity", bins = 50, alpha = 0.5)
```

```{r}
valid_draws_df |> 
  ggplot() +
  aes(x = deltaTheta/pi) +
  geom_histogram() +
  geom_vline(xintercept = with(input_params, theta[2] - theta[1]), colour = "red")
```

```{r}
valid_draws_df |> 
  ggplot() +
  aes(x = `theta[1]`, y = `theta[2]`) +
  geom_point()
```

```{r eval=FALSE}
valid_draws_df |> 
  ggplot() +
  aes(x = `phi[1]`, y = `phi[2]`) +
  geom_point()
```

```{r eval=FALSE}
valid_draws_df |> 
  ggplot() +
  aes(x = `phi[1]`, y = `theta[1]`) +
  geom_point()
```

```{r eval=FALSE}
valid_draws_df |> 
  ggplot() +
  aes(x = `phi[2]`, y = `theta[2]`) +
  geom_point()
```
